How Slow do my Transient Typechecks Go?

### 2019-08-17 paper.tex
Correct abbreviation of everyone but Isaac’s name
Signed-off-by: Stefan Marr <git@stefan-marr.de>
Edit intro to section 3, incl. sec. 3.1
Signed-off-by: Stefan Marr <git@stefan-marr.de>

\author[Richard R.]{Richard Roberts}
\affiliation{
  \department{Computational Media Innovation Centre} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{rykardo.r@gmail.com} % \email is recommended

\author[Stefan M.]{Stefan Marr}
\orcid{0000-0001-9059-5180}
\affiliation{
  \department{School of Computing} % \department is recommended
  \institution{University of Kent}
  \country{United Kingdom}
}
\email{s.marr@kent.ac.uk} % \email is recommended

\author[Michael H.]{Michael Homer}
\affiliation{
  \department{Engineering and Computer Science} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{mwh@ecs.vuw.ac.nz} % \email is recommended

\author[James N.]{James Noble}
\orcid{0000-0001-9036-5692}             %% \orcid is optional
\affiliation{
  \department{Engineering and Computer Science} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{kjx@ecs.vuw.ac.nz} % \email is recommended

@@

\begin{itemize}
  \item an approach to identify gradual type annotations
        that cause significant performance effects
  \item adapting the Takikawa protocol for languages and implementations
        where performance effects are dominated by type annotations instead
        of transitions between typed and untyped modules
  \item observing that performance of small benchmarks
        with 10--250 type annotations is dominated by only
        one or two type annotations
\end{itemize}

@@

Moth relies on a number of standard techniques for optimising
object-oriented programs.
``Shapes'' \citep{woss2014object} capture information about objects'
structures and (run time) 
field types, allowing a just-in-time compiler to
represent objects in memory similarly to C structs and, consequently,
can generate highly efficient code.
``Polymorphic inline caches''
\citep{Hoelzle:91:PIC} use object shapes to cache the results of
method lookups, avoiding expensive class hierarchy searches or
indirect jumps through virtual method tables. 
Since Moth is built on the Truffle framework,
Graal comes with  additional support for partial evaluation,
which enables efficient native code generation for
Truffle interpreters \citep{Wurthinger:2017:PPE}.
\RR{This last sentence seems out of place, does it belong here?}
\kjx{yes I think it's fine}

\section{Experimental Methodology}
Our methodology is built upon the so-called ``Takikawa'' or ``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp} benchmarks a \emph{lattice} of $2^N$ \emph{configurations} of each benchmark, where a configuration is a particular mix of static and dynamically typed code, however here we only test a relatively small sample of this lattice. Here we apply this methodology to our previous work \cite{roberts-and-co-ecoop-2019}, where we evaluated the performance of transient type checks in Moth by using a suite of benchmarks.

@@

To investigate Moth's support for transient type checks further, we carried out a similar approach to the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions (fewer if the benchmark was too small). We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

\subsection{The Benchmarks}
In our previous work \cite{roberts-and-co-ecoop-2019} we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks. We added another nine benchmarks taken from the gradual-typing literature.
Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
We surveyed a number of papers
\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected additional benchmarks that have been used by multiple papers.
Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviourally equivalent,
we chose the \AWFY versions since we already used them to
establish a performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
Table~\ref{tab:gradual-benchmarks}.

>>>>

\author[R. Roberts]{Richard Roberts}
\affiliation{
  \department{Computational Media Innovation Centre} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{rykardo.r@gmail.com} % \email is recommended

\author[S. Marr]{Stefan Marr}
\orcid{0000-0001-9059-5180}
\affiliation{
  \department{School of Computing} % \department is recommended
  \institution{University of Kent}
  \country{United Kingdom}
}
\email{s.marr@kent.ac.uk} % \email is recommended

\author[M. Homer]{Michael Homer}
\affiliation{
  \department{Engineering and Computer Science} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{mwh@ecs.vuw.ac.nz} % \email is recommended

\author[J. Noble]{James Noble}
\orcid{0000-0001-9036-5692}             %% \orcid is optional
\affiliation{
  \department{Engineering and Computer Science} % \department is recommended
  \institution{Victoria University of Wellington}
  \country{New Zealand}
}
\email{kjx@ecs.vuw.ac.nz} % \email is recommended

@@

\begin{itemize}
  \item an approach to identify gradual type annotations
        that cause significant performance effects
  % SM: after discussion with Isaac, this was done
  %     by some Reticulated Python paper before
  % and Isaac's adaptation is restricted to using rounding rules
  %    (>= 0.5 is round up) that give useful results (at least 1 type check)
  %    for selecting the number of type checks
  % \item adapting the Takikawa protocol for languages and implementations
  %       where performance effects are dominated by type annotations instead
  %       of transitions between typed and untyped modules
  \item observing that performance of small benchmarks
        with 10--250 type annotations is dominated by only
        one or two type annotations
\end{itemize}

@@

Moth relies on a number of standard techniques for optimising
object-oriented programs.
``Shapes'' \citep{woss2014object} capture information about objects'
structures and (run time) 
field types, allowing a just-in-time compiler to
represent objects in memory similarly to C structs and, consequently,
can generate highly efficient code.
``Polymorphic inline caches''
\citep{Hoelzle:91:PIC} use object shapes to cache the results of
method lookups, avoiding expensive class hierarchy searches or
indirect jumps through virtual method tables. 
Since Moth is built on the Truffle framework,
Graal comes with  additional support for partial evaluation,
which enables efficient native code generation for
Truffle interpreters \citep{Wurthinger:2017:PPE}.
% SM: kjx says it's fine, so, drop for now
% \RR{This last sentence seems out of place, does it belong here?}
% \kjx{yes I think it's fine}

\section{Experimental Methodology}

Our goal is to identify which type annotations in Grace programs
cause performance effects.
To this end, we built upon the so-called ``Takikawa'' or ``Takikawa-Greenman'' evaluation protocol \cite{Takikawa2016,Greenman2019jfp}.
It uses $2^N$ \emph{configurations} of each benchmark.
A configuration is a particular mix of static and dynamically typed code, forming a lattice of configurations.
We only test a relatively small sample of this lattice,
which is in our experience sufficient to pinpoint performance anomalies caused
by type annotations.

@@

 % Here we apply this methodology to our previous work \cite{roberts-and-co-ecoop-2019}, where we evaluated the performance of transient type checks in Moth by using a suite of benchmarks.

To identify type annotations causing anomalies,
% To investigate Moth's support for transient type checks further, 
we adapted the Takikawa-Vitousek protocol. For each benchmark we generated 100 partially typed versions,
or fewer if the benchmark has at most 10 types.
We did an even split so that for each interval $i \ge 1$ and $i$ < $N$, we generated roughly the same number of configurations with $i$ type annotations. We used Robert Floyd's sampling algorithm \cite{Bentley:1987:PPS:30401.315746} to choose the type annotations each configuration contained randomly, and we ensured that no duplicate configurations were generated. In addition to these, we tested fully untyped and typed versions, for a total of 102 configurations per benchmark (or 97 in the case of our Storage benchmark, since it only has 10 type annotations).

\subsection{The Benchmarks}
In our previous work \cite{roberts-and-co-ecoop-2019}, we evaluated the performance of transient type checks using 21 benchmarks. We started with the 14 benchmarks of the \AWFY
benchmark suite \cite{Marr2016}: Bounce, CD, DeltaBlue, GraphSearch, Havlak, Json, List, Mandelbrot, NBody, Permute, Richards, Sieve, Storage and Towers benchmarks. We added another nine benchmarks taken from the gradual-typing literature.
Our goal was to complement the \AWFY with additional ones that are used for similar experiments and can be ported to Grace.
We surveyed a number of papers
\citep{Takikawa2016,Vitousek2017,Muehlboeck2017,Bauman2017,Richards2017,Stulova2016,Greenman2018}
and selected additional benchmarks that have been used by multiple papers.
Two of these benchmarks (NBody and Sieve) overlapped with the \AWFY suite,
or were available in different versions.
While not always behaviourally equivalent,
we chose the \AWFY versions since we already used them to
establish a performance baseline.
The selected benchmarks as well as the papers in which they were used are shown in
Table~\ref{tab:gradual-benchmarks}.
